\PYGEOMETRY{}: a Python library for the creation of Monte Carlo radiation transport physical geometries

### 2020-09-05 public/paper/pyg4ometry.tex
Paper update

\newcommand{\pyinline}[1]{\lstinline[postbreak={}]{#1}}

@@

% part features
% part assembly

@@

%%%%%%%%%%%%%%%%% Geant4 geometery
Geant4 geometry description is a mixture of BREP, CSG and tessellated concepts. Geant4 includes 27 basic solids, although it does not store a sense 
of topology present in traditional CAD BREP systems. One of the fundamental solids is a {\em TessellatedSolid} which can used to represent STL or PLY files. 
Geant4 also provides the ability to perform Boolean operations on these primitive solids. The richest and most flexible geometry description is currently used by
Geant4. Not only do solid objects need to be defined but also placed in a world coordinate system. Geant4 has two concepts which facilitate this {\it Logical Volumes} and {\it Physical Volumes}. A {\it Logical Volume} is a region of space  that is defined by an outer solid but also other attributes like material, magnetic field 
and zero or more physical {\it daughter} volumes. A Physical Volume is a placement (or instance) of a Logical Volume.

@@

%%%%%%%%%%%%%%%%%  Input and output  
For each input format available in \PYGEOMETRY{}  (GDML, STL, FLUKA and STEP) a dedicated {\it Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and provides a \verb|Registry| which can be used or manipulated by the user. 
Output consists of taking the registry and writing the appropriate GDML.

@@

%%%%%%%%%%%%%%%%%  Replica, division, parametrised and looped volumes
A powerful feature of Geant4 and hence GDML is the ability to either repeat, divide or parametrise geometry. The class which enables the creation of 
multiple replicas of a volume in a Cartesian, cylindrical or spherical grid is known as a Replica Volume. A Division Volume breaks a primitive into segments 
in either a Cartesian or cylindrical polar coordinates. A Parametrised Volume allows for the arbitrary multiple placement of solids where the parameters are 
allowed to vary for each placement.  Another way to in GDML to create parametrised solids or volumes is GDML loops, where sections of GDML are allowed 
to be repeated with varying parameters. GDML loops loading and expansion are not supported by \PYGEOMETRY{} and will be implemented in a 
future release.

@@

%%%%%%%%%%%%%%%%% Volume bounding
If a user is creating a placing multiple {\em daughter} volumes within a {\em mother} volume then it is the users responsibility to create a solid which fully 
encompasses the daughter volumes. Overlaps between daughter volumes and the mother can be detected, but it is desirable to have a mother volume 
shape that efficiently holds all of its daughters.

@@

The {\it Physical Volume} class (\pyinline{pyg4ometry.geant4.PhysicalVolume}) is also used to store visualisation attributes like the solid's 
colour, surface or wireframe representation and visibility. Overlaps detected in the mesh geometry are
stored in the Logical Volume and can be  displayed separately to allow a user to understand the reason for 
the overlaps.

@@

Each FLUKA body is represented in \PYGEOMETRY{} with a corresponding
class, and in turn each class has methods which returns a GDML
primitive solid and that solid's rotation and position such that it
matches its FLUKA equivalent.  The expansion, translation and
transform geometry directives are each folded into one or more of
these three methods.  The mapping of FLUKA bodies to GDML solids is
shown in Table~\ref{tab:Fluka2Geant4}.  It is worth noting that many
of the FLUKA bodies are infinite in extent, but are mapped to finite
GDML solids.  The translation of infinite bodies to equivalent finite
solids is one of the main steps involved in the conversion process.
This mapping is possible because whilst FLUKA bodies can be infinite
in extent, all zones and regions must be finite.  Zones and regions
are then composed by instantiating these classes and adding body
instances to them.  Each Zone and Region instance can then return its
equivalent GDML Boolean solid.

@@

\subsection{Materials}
\PYGEOMETRY{} converts GDML/Geant4 materials to FLUKA \verb|MATERIAL| and \verb|COMPOUND| cards.

\subsection{Disucssion}
There are still a few outstanding issues with Geant4/GDML to FLUKA conversion.

\begin{enumerate}
\item Degenerate body look up 
\item Bodies with/without transformation 
\item Non-convex solids, convex decomposition (TwistedBox, TwistedTrap, TwistedTrd, TwistedTube, Tessellated)
\item Scaled solid, Placement scale 
\item Replica, Division, Parameterised
\item Recursive application of \ref{eqn:setDiff}, \ref{eqn:setUnion}, \ref{eqn:setIntersection}, \ref{eqn:logicalSubtraction} speed up using meshes
\item Issues with tessellated solids 
\end{enumerate}

@@

@@

\section{Conclusions}
The authors believe that tools to quickly create, either ab-initio or by conversion geometry usable in 
particle transport Monte Carlo programmes will save significant amounts of user effort and ultimately 
yield more accurate simulations. \PYGEOMETRY{} is a relatively complete implementation of a geometry 
creation tool, whilst heavily internally based on Geant4 and GDML can have utility for users of all RT 
codes. \PYGEOMETRY{} can clearly be extended to other formats or applications. As for applications, examples 
include  the  generation triangular mesh structures for GPU accelerated photon tracking in liquid noble dark 
matter detectors and geometries for event visualisation. Although not implemented in the current version, 
a C++ output writer can be quickly implemented to generate geometry which can be compiled into a Geant4 
application.

>>>>

\newcommand{\pyinline}[1]{\lstinline[postbreak={}]{#1}}
\newcommand{\cpinline}[1]{\lstinline[postbreak={}]{#1}}

@@

% part features
% part assembly
% physical volume
% logical volume
% daughter volume
% mother volume
% Reader
% scene

@@

%%%%%%%%%%%%%%%%% Geant4 geometery
Geant4 geometry description is a mixture of BREP, CSG and tessellated concepts. Geant4 includes 27 basic solids, although it does not store a sense 
of topology present in traditional CAD BREP systems. One of the fundamental solids is a tessellated sollid which can used to represent STL or PLY files. 
Geant4 also provides the ability to perform Boolean operations on these primitive solids. The richest and most flexible geometry description is currently used by
Geant4. Not only do solid objects need to be defined but also placed in a world coordinate system. Geant4 has two concepts which facilitate this {\em logical volumes} and {\em physical volumes}. A logical volume is a region of space that is defined by an outer solid but also other attributes like material, magnetic field 
and zero or more physical {\em daughter} physical volumes. A physical volume is a placement (or instance) of a logical volume.

@@

%%%%%%%%%%%%%%%%%  Input and output  
For each input format available in \PYGEOMETRY{}  (GDML, STL, FLUKA and STEP) a dedicated {\em Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and provides a \verb|Registry| which can be used or manipulated by the user. 
Output consists of taking the registry and writing the appropriate GDML.

@@

%%%%%%%%%%%%%%%%%  Replica, division, parametrised and looped volumes
A powerful feature of Geant4 and hence GDML is the ability to either repeat, divide or parametrise geometry. The class which enables the creation of 
multiple replicas of a volume in a Cartesian, cylindrical or spherical grid is known as a Replica Volume. A Division Volume breaks a primitive into segments 
in either a Cartesian or cylindrical polar coordinates. A parametrised volume allows for the arbitrary multiple placement of solids where the parameters are 
allowed to vary for each placement.  Another way to in GDML to create parametrised solids or volumes is GDML loops, where sections of GDML are allowed 
to be repeated with varying parameters. GDML loops loading and expansion are not supported by \PYGEOMETRY{} and will be implemented in a 
future release.

@@

%%%%%%%%%%%%%%%%% Volume bounding
If a user is creating a placing multiple daughter volumes within a {\em mother} volume then it is the users responsibility to create a solid which fully 
encompasses the daughter volumes. Overlaps between daughter volumes and the mother can be detected, but it is desirable to have a mother volume 
shape that efficiently holds all of its daughters.

@@

The physical volume class (\pyinline{pyg4ometry.geant4.PhysicalVolume}) is also used to store visualisation attributes like the solid's 
colour, surface or wireframe representation and visibility. Overlaps detected in the mesh geometry are
stored in the Logical Volume and can be  displayed separately to allow a user to understand the reason for 
the overlaps.

@@

Each FLUKA body is represented in \PYGEOMETRY{} with a corresponding
class, and in turn each class has methods which returns a GDML
primitive solid and that solid's rotation and position such that it
matches its FLUKA equivalent.  The expansion, translation and
transform geometry directives are each folded into one or more of
these three methods.  The mapping of FLUKA bodies to GDML solids is
shown in Table~\ref{tab:Fluka2Geant4}.  It is worth noting that many
of the FLUKA bodies are infinite in extent, but are mapped to finite
GDML solids.  The translation of infinite bodies to equivalent finite
solids is one of the main steps involved in the conversion process.
This mapping is possible because whilst FLUKA bodies can be infinite
in extent, all zones and regions must be finite.  Zones and regions
are then composed by instantiating these classes and adding body
instances to them.  Each zone and region instance can then return its
equivalent GDML Boolean solid.

@@

\subsection{Materials}
\PYGEOMETRY{} converts GDML/Geant4 materials to FLUKA \verb|MATERIAL| and \verb|COMPOUND| cards.
Geant4 has a class \cpinline{G4Material} to assign material state (density, physical state, temperature and pressure) 
to a logical volume. \cpinline{G4Material} has two main constructors, the first were a atomic number is supplied and 
the second is when \cpinline{G4Element} instances and relative atomic or mass abundances are provided. The {\em simple}
\cpinline{G4Material} is converted to a \fluka{material} card, whilst the {\em element} \cpinline{G4Material}  is converted to a 
\fluka{compound} card. A similar issues existing when converting \cpinline{G4Element} to FLUKA, as \cpinline{G4Element}
can either be simple, so defined only by atomic number and mass or composite and defined by a admixture defined by 
relative abundances of \cpinline{G4Isotope}. A similar mapping is performed so if an \cpinline{G4Element} is simple it is 
directly converted to a FLUKA \fluka{MATERIAL} card, and the {\it isotope} \cpinline{G4Element} is converted to a \fluka{compund}
card.

\subsection{Disucssion}
Overall the conversion to FLUKA input format from GDML is quite 
advanced and stable. Relatively large experimental simulations 
have been converted from GDML to FLUKA and have been producing 
simulation results. The conversion described still requires a user to
understand how geometry is specified in both Geant4 and FLUKA. For 
example the subtraction of many non-primitive body daughter volumes from 
a mother will create a very complex non-DNF region which is inefficient when viewed in 
FLAIR and used for simulation in FLUKA. A geometry designer can avoid this by 
restricting daughter volumes in Geant4 that can be expressed without 
XXX. If \PYGEOMETRY{} is used during the creation of geometry then multiple 
codes can be targeted without additional user effort. There are still a few outstanding technical 
issues, which should be discussed.

Currently replica, division and parametrised placements are not implemented 
and will be addedin a future release. In Geant4 it is possible to create scaled 
solids or placements with reflections, so called scale in Geant4. FLUKA 
roto-translations do not support reflections and implementing reflections 
require transformation of the body definitions. This is not yet implemented 
in the current version of \PYGEOMETRY{}.

It is possible given the GDML to FLUKA conversion algorithm described 
in this paper that coplanar overlaps exist in the FLUKA geometry.  In 
Geant4 there is no connection between surfaces used to specify on 
logical volume solid and another logical volume solid. So for example 
if a logical volume solid shared a face with one its daughter volume 
solids the body would be duplicated in the final FLUKA file. It 
is possible to remove obvious degeneracies but this is complicated by 
placements of bodies. It should be possible to cast FLUKA bodies 
{\em without} roto-translations into a {\em normal form} which can be used 
to test for approximate equality. Approximate equality is required as 
multiple application roto-translations will accrue numerical rounding errors.
An simple example of this is the XYP and PLA, it simple to transform an 
XYP into a PLA with an appropriate rotation. This would allow for the removal 
of degenerate surfaces removing potential coplanar overlaps and also reduce 
the final converted file size.

The twisted primitives need to be decomposed 
into a union of convex solids. This decomposition does not always 
succeed or produces a far from optimal number of convex solids. 
An alternative to implement these solids is to approximate each 
layer of the twisted solid as a union of terahedra. A similar problem 
exists for general tessellated solids, which in general
are non-convex and need decomposition into convex hulls. A potential 
way to avoid a computationally expensive decomposition step is to 
create a region formed from the unions of tetrahedra. There are numerous 
algorithms for tetrahedralisation of surface meshes in both CGAL and TetGen 
and will be implemented in a future release. Even if a stable and general 
method for converting tessellated solids exists it not efficient to define 
tessellated objects in this way and memory, body or zone limits might
be reached in FLUKA and so limit the size of CAD or STL models which 
might be loaded.

\begin{enumerate}
%\item Degenerate body look up 
%\item Bodies with/without transformation 
% \item Non-convex solids, convex decomposition (TwistedBox, TwistedTrap, TwistedTrd, TwistedTube, Tessellated)
% \item Scaled solid, Placement scale 
% \item Replica, Division, Parameterised
\item Recursive application of \ref{eqn:setDiff}, \ref{eqn:setUnion}, \ref{eqn:setIntersection}, \ref{eqn:logicalSubtraction} speed up using meshes
% \item Issues with tessellated solids 
\end{enumerate}

@@

\begin{enumerate}
\item Mesh overlap and shrinking algorithm 
\item Surface conversion and decomposition algorithm (future release)
\end{enumerate}

@@

\section{Conclusions and discussion}
The authors believe that tools to quickly create, either ab-initio or by conversion geometry usable in 
particle transport Monte Carlo programmes will save significant amounts of user effort and ultimately 
yield more accurate simulations. \PYGEOMETRY{} is a relatively complete implementation of a geometry 
creation tool, whilst heavily internally based on Geant4 and GDML can have utility for users of all RT 
codes. \PYGEOMETRY{} can clearly be extended to other formats or applications. As for applications, examples 
include  the  generation triangular mesh structures for GPU accelerated photon tracking in liquid noble dark 
matter detectors and geometries for event visualisation. Although not implemented in the current version, 
a C++ output writer can be quickly implemented to generate geometry which can be compiled into a Geant4 
application.
